\chapter{Data Input}\index{general}{data input}

This chapter deals with how to get read in data for statistical analysis (in \emph{Python}), and thus forms the link between the chapter on \emph{Python} and the first chapter on data analysis. And while it may be hard to believe, reading data into the system in the correct format is often one of the most time consuming parts of the data analysis.

\vspace{5 mm}

Data input can be complicated by a number of problems, like different separators between data entries (such as spaces and/or tabulators), or empty lines at the end of the file. In addition, data may have been saved in different formats, such as MS Excel, \emph{Matlab},  HDF5 (which also contains the \emph{Matlab}-format), or in databases. Understandably, we cannot cover all possible combinations here. But I will try to give an overview of where and how to start with data input.

\section{Data from Textfiles}

\subsection{Visual Inspection}

When your data are available as ASCII-files, you should always start out with a visual inspection of the data! In particular, you should check

\begin{itemize}
  \item Do the data have a header and/or a footer?
  \item Are there empty lines at the end of the file?
  \item Are there white-spaces before the first number, or at the end of each line? (The latter is a lot harder to see.)
  \item Are the data separated by tabulators, and/or by spaces? (Tip: you should use a text-editor which can visualize tabs, spaces, and end-of-line (EOL) characters.)
\end{itemize}

\subsection{Reading ASCII-data into Python}

In \emph{Python}, I strongly suggest that you start out reading in and inspecting your data in the \emph{ipython qtconsole}. It allows you to move around much more easily, try things out, and quickly get feedback on how successful your commands were. When you have your command syntax worked out, you can obtain the command history with \lstinline{%history}, copy it into your IDE, and turn it into a program.

While the a numpy command \lstinline{np.loadtxt} allows you to read in simply formatted text data, most of the time I go straight to \emph{pandas}, as it provides significantly more powerful tools for data-entry. A typical workflow can contain the following lines:

\begin{lstlisting}
    In [1]: import pandas as pds
    In [2]: cd 'C:\Data\storage'
    In [3]: pwd     # Check if you were successful: this is not always the case!
    In [4]: ls      # List the files in that directory
    In [5]: inFile = 'data.txt'
    In [6]: df = pd.read_csv(inFile)
    In [7]: df.head()   # Check if the first line was read in properly, and compare it to the values in the ASCII-file
    In [8]: df.tail()   # Check the last line, and compare it to the values in the ASCII-file
\end{lstlisting}

After "In [7]" I typically have to play around with the options of \lstinline{pd.read_csv}, to get things right.
Make sure you check that the number of column headers is equal to the number of columns that you expect. It can happen that everything gets red in - but into one large single column!

\subsubsection{Simple Text-file}

For example, to read in a file "data.txt" with the following content

\begin{lstlisting}
    1, 1.3, 0.6
    2, 2.1, 0.7
    3, 4.8, 0.8
    4, 3.3, 0.9
\end{lstlisting}

could be read in and displayed with

\begin{lstlisting}
    In [16]: data = np.loadtxt('data.txt', delimiter=',')

    In [16]: data
    Out[16]:
        array([[ 1. ,  1.3,  0.6],
               [ 2. ,  2.1,  0.7],
               [ 3. ,  4.8,  0.8],
               [ 4. ,  3.3,  0.9]])
\end{lstlisting}

where \lstinline{data} is a \emph{numpy array}. Without the flag \lstinline{delimiter=','}, the function \emph{np.loadtxt} crashes. An alternative way to read in these data is with

\begin{lstlisting}
  In [17]: df = pd.read_csv('data.txt', header=None)
  
  In [18]: df
  Out[18]:
       0    1    2
    0  1  1.3  0.6
    1  2  2.1  0.7
    2  3  4.8  0.8
    3  4  3.3  0.9
\end{lstlisting}

where \lstinline{df} is a \emph{pandas DataFrame}. Without the flag \lstinline{header=None}, the first row is falsely interpreted as the column labels!

\begin{lstlisting}
  In [19]: df = pd.read_csv('data.txt')

  In [20]: df
  Out[20]:
       1   1.3   0.6
    0  2   2.1   0.7
    1  3   4.8   0.8
    2  4   3.3   0.9
\end{lstlisting}
 
 The \emph{pandas} routine has the advantage that the first column is recognized as integer, whereas the second and third column are float.

\subsubsection{More complex Text-files}

The advantage of using \emph{pandas} for data input becomes clear with more complex files. Take for example the following input file "data2.txt":

\begin{lstlisting}
    ID, Weight, Value
    1, 1.3, 0.6
    2, 2.1, 0.7
    3, 4.8, 0.8
    4, 3.3, 0.9

    Those are dummy values, created by ThH.
    June, 2015
\end{lstlisting}

One of the input flags of \lstinline{pd.read_csv} is \lstinline{skipfooter}, so we can read in the data easily with

\begin{lstlisting}
  In [21]: df2 = pd.read_csv('data.txt', skipfooter=3, delimiter='[ ,]*')
\end{lstlisting}

The last option, \lstinline{delimiter='[ ,]*'}, is a "regular expression" (see below) specifying that "one or more spaces and/or commas may be used to seperate entry values". Also, when the input file includes a row with the column names,  the data can be accessed immediately with their corresponding column name, e.g.:

\begin{lstlisting}
  In [22]: df2
  Out[22]:
       ID   Weight   Value
    0   1      1.3     0.6
    1   2      2.1     0.7
    2   3      4.8     0.8
    3   4      3.3     0.9

    In [23]: df2.Value
    Out[23]:
    0    0.6
    1    0.7
    2    0.8
    3    0.9
    Name: Value, dtype: float64
\end{lstlisting}

\subsection{Regular Expressions}\index{general}{regular expressions}

Working with text data often requires the use of simple \emph{regular expressions}\index{general}{regular expressions}. \Glspl{regexp} are a very powerful way of finding and/or manipulating text strings. Many books have been written about them.

\begin{itemize}
  \item \url{https://www.debuggex.com/cheatsheet/regex/python} provides a convenient cheat sheet for regular expressions in \emph{Python}.
  \item \url{http://www.regular-expressions.info} gives and a comprehensive description.
\end{itemize}

Let me give you two useful examples:

\begin{enumerate}
  \item \lstinline{df = pd.read_csv(inFile, sep='[ ;,]*'}

  would read in data that are separated by a \emph{combination} ("[...]") of \emph{one or more} ("*") whitespaces, semicolons, and/or commas.

  \item \lstinline{vel = df.filter(regex='Vel*')}

  would extract all the columns that start with "Vel...". For example, if your DataFrame \emph{df} contains for example the columns \emph{['Time', 'PosX', 'PosY', 'PosZ', 'VelX', 'VelY', 'VelZ']}, this command would return a DataFrame with only the columns \emph{['VelX', 'VelY', 'VelZ']}.
\end{enumerate}

\section{Input from MS Excel}\index{general}{Excel}

\emph{pandas} also offers tools for reading and writing data for \emph{Microsoft Excel}, in-memory data structures and SQL databases.

\footnote{The following section has been taken from the \emph{pandas} documentation.}There are two approaches to reading an \emph{Excel} file. The \lstinline{read_excel} function and the \lstinline{ExcelFile} class. \lstinline{read_excel} is for reading one file with file-specific arguments (ie. identical data formats across sheets). \lstinline{ExcelFile} is for reading one file with sheet-specific arguments (ie. various data formats across sheets). Choosing the approach is largely a question of code readability and execution speed.

Equivalent class and function approaches to read a single sheet:

\begin{lstlisting}
    # using the ExcelFile class
    xls = pd.ExcelFile('path_to_file.xls')
    data = xls.parse('Sheet1', index_col=None, na_values=['NA'])

    # using the read_excel function
    data = read_excel('path_to_file.xls', 'Sheet1', index_col=None, na_values=['NA'])
\end{lstlisting}

If this fails, give it a try with the \emph{Python} package \emph{xlrd}.

\textbf{Example: }
\PyImg "readZip.py" (p \pageref{py:readZip}) This advanced script shows you how you can directly import data from an \emph{Excel} file from a zipped archive on the web.
\index{python}{readZip}

\section{Input from Matlab and other formats}

\begin{description}
  \item[Matlab] \emph{Matlab} support is built into \emph{scipy}, with the command \lstinline{scipy.io.loadmat}.
  \item[Clipboard] If you have data in your clipboard, you can import them directly with \lstinline{pd.read_clipboard()}
  \item[Other Fileformats] pandas supports input from a number of additional formats. The simplest way to access them is typing \lstinline{pd.read_ + TAB}, which shows you the currently available options.
\end{description}

\subsection{Matlab}
The following commands return string, number, vector, and matrix variables from a \emph{Matlab} file "data.mat", as well as the content of a structure with two entries (a vector and a string):

\begin{lstlisting}
    from scipy.io import loadmat
    data = loadmat('data.mat')
    
    number = data['number'][0,0]
    text   = data['text'][0]
    vector = data['vector'][0]
    matrix = data['matrix']
    struct_values = data['structure'][0,0][0][0]
    strunct_string = data['structure'][0,0][1][0]
\end{lstlisting}

