\makeglossaries

\newglossaryentry{function}
{
    name=function,
    description={A Python object that accepts input data, executes commands and
    calculations with them, and can return one or more return objects.}
}

\newglossaryentry{module}
{
    name=module,
    description={A file containing variables and function definitions.}
}

\newglossaryentry{package}
{
    name=Python package,
    description={A folder containing one or more Modules and an ".ini"-file}
}

\newglossaryentry{IPython}
{
    name=IPython,
    description={The IPython kernel is an interactive shell that allows the immediate
    execution of Python commands, and extends these with "magic functions"
    (e.g. "%cd", or "%whos") which facilitat the interactive work. It can be
    executed in a terminal modus, in a "qtconsole" which allows the display of
    help and the graphical outputs in the command window, and in a "notebook",
    a browser-based implementation with support for code, rich text,
    mathematical expressions, inline plots and other rich media. The
    "notebook" has been so successful that it has been applied to other
    languages, and has now become an independent project called "Jupyter".}
}

\newglossaryentry{pylab}
{
    name=pylab,
    description={pylab is a convenience module that bulk imports matplotlib.pyplot (for
    plotting) and numpy (for mathematics and working with arrays) in a single
    name space.}
}

\newglossaryentry{numpy}
{
    name=numpy,
    description={ Python package for numerical data manipulation. Provides efficient
    handling of mathematical manipulations on vectors and on arrays of two or
    more dimensions.}
}

\newglossaryentry{scipy}
{
    name=scipy,
    description={ Python package for scientific computation. Based on numpy.}
}

\newglossaryentry{regexp}
{
    name=regular expression,
    description={A sequence of characters that define a search pattern, mainly for use in
    pattern matching with strings. Available for Unix, Python, Perl, Java, C++, and many other
    languages.}
}

\newglossaryentry{categorical}
{
    name=categorical data,
    description={ Data that can take on one of a limited, and usually fixed, number of
    possible values, with no natural order. (If a "mean value" makes no
    sense.)}
}

\newglossaryentry{numerical}
{
    name=numerical data,
    description={ Data that can be expressed by a (continuous or discrete)
    number.}
}

\newglossaryentry{density}
{
    name=density,
    description={ Density (as in Kernel-density or Probability-density): A continuous function that describes the relative likelihood for this
    random variable to take on a given value.}
}

\newglossaryentry{distribution}
{
        name=distribution,
        description={A function which assigns a probability to each measurable subset of the possible outcomes of a random experiment.}
}

\newglossaryentry{boxplot}
{
    name=box-plot,
    description={A common visualisation of the distribution of data, expressed by a box
    with a line inside that box, and whiskers at the top and bottom.
    The box indicates the first and third quartile, and the line the median
    value of the data sample. The whiskers can indicate either the range of
    the data, or the most extreme value within 1.5*the inner-quartile-range.}
}

\newglossaryentry{population}
{
    name=population,
    description={ Includes all of the elements from a set of data.}
}

\newglossaryentry{sample}
{
    name=sample,
    description={ One or more observations from a population.}
}

\newglossaryentry{observational}
{
    name=observational study,
    description={ Study where the assignment of subjects into a treated group versus a
    control group is outisde the control of the investigator.}
}

\newglossaryentry{experimental}
{
    name=experimental study,
    description={ Study where the selection of the subjects as well as the conditions of the
    study are under the control of the investigator.}
}

\newglossaryentry{prospective}
{
    name=prospective study,
    description={ A prospective study watches for outcomes, such as the development of a
    disease, during the study period and relates this to other factors such as
    suspected risk or protection factor(s).}
}

\newglossaryentry{retrospective}
{
    name=retrospective study,
    description={ A retrospective study looks backwards and examines exposures to suspected
    risk or protection factors in relation to an outcome that is established
    at the start of the study. }
}

\newglossaryentry{treatment}
{
    name=treatment,
    description={ Same as factor.}
}

\newglossaryentry{factor}
{
    name=factor,
    description={ Also called "treatment" or "independent variable", is an explanatory
    variable manipulated by the experimenter.}
}

\newglossaryentry{mode}
{
    name=mode value,
    description={ The highest value value in a discrete or continusoud
    probability distribution.}
}

\newglossaryentry{median}
{
    name=median value,
    description={ The value separating the higher half of the data sample form the lower
    half.}
}

\newglossaryentry{casecontrol}
{
    name=case control study,
    description={An a type of observational study in which two existing groups differing in
    outcome are identified and compared on the basis of some supposed causal
    attribute. "First treat, then select."}
}

\newglossaryentry{cohort}
{
    name=cohort study,
    description={ A type of observational study, where you first select the patients, and
    then follow their deveopment. For instance in medicine, it is an analysis
    of risk factors and follows a group of people who do not have the disease,
    and uses correlations to determine the absolute risk of subject
    contraction.}
}

\newglossaryentry{crossover}
{
    name=crossover study,
    description={A longitudinal study in which all subjects receive a sequence of different
    treatments.}
}

\newglossaryentry{randomization}
{
    name=randomization,
    description={ A method to eliminate bias from research results, by dividing a
    homogeneous group of subjects into a "control group" (which receives no
    treatment) and a treatment group.}
}

\newglossaryentry{block}
{
    name=block,
    description={To reduce the variability from a variable that cannot be randomized by
    blocking/fixating it.}
}

\newglossaryentry{bias}
{
    name=bias,
    description={ Systematic deviation from a sample statistic from the corresponding
    population statistic. Often caused by uncareful selection of subjects.}
}

\newglossaryentry{location}
{
    name=location,
    description={ Paramter that shifts the mean of a probability
    distribution.}
}

\newglossaryentry{scale}
{
    name=scale,
    description={ Parameter that controls the variance of a probability
    distribution.}
}

\newglossaryentry{shapeparameter}
{
    name=shape parameter,
    description={ Parameters beyond "location" and "scale" which control the shape of a
    probability distribution. Rarely used.}
}

\newglossaryentry{standarderror}
{
    name=standard error,
    description={ Often short for "standard error of the mean". Square root of the variance
    of a statistic.}
}

\newglossaryentry{confidenceinterval}
{
    name=confidence interval,
    description={ Interval estimate of a population parameter, which contains the true value
    of the parameter with a defined percent-likelihood (e.g. 95\%-CI).}
}

\newglossaryentry{quantile}
{
    name=quantile,
    description={ Value that is larger or equal than p*100% of the data, where
    0<p<=1}
}

\newglossaryentry{quartile}
{
    name=quartile,
    description={ Value that is larger or equal than 25% / 50% / 75% of the data (first /
    second / third quartile)}
}

\newglossaryentry{percentile}
{
    name=percentile,
    description={ Value that is larger or equal than p% of the data, where 1<=
    p < 100.}
}

\newglossaryentry{rank}
{
    name=rank,
    description={Ranked data are numbered data, where the number corresponds to the rank of the data
    (number in the sequence of the sorted data), and not to a
    continuous value.}
}

\newglossaryentry{matplotlib}
{
        name=Matplotlib,
        description={A Python package which provides the ability to generate 2D and 3D plots. This includes the plotting commands, as well as the functionality of different output media, also called \emph{backends}: an output can for example be into the IPython notebook; or into a PDF-file; or into a separate graphics window.}
}

\newglossaryentry{poweranalysis}
{
        name=power analysis,
        description={Calculation of the minimum sample size required so that one can be reasonably likely to detect an effect of a given size.}
}

\newglossaryentry{pdf_}
{
    name= probability density function,
    description={ Defines the likelihood to find a random variable to obtain a certain
    value. The probability to find the value of the random variable in a given
    interval is the integral of the probability density function over that
    interval.}
}

\newglossaryentry{cdf_}
{
    name=cumulative distribution function,
    description={ The probability to find a random variable with a value lower than the
    given one.}
}

\newglossaryentry{sd_}
{
    name=standard deviation,
    description={ Square root of variance}
}

\newglossaryentry{variance}
{
    name=variance,
    description={ Measure of how far a set of numbers is spread out. Mathematically, it is
    the expected value of the squared deviation from the mean:
    $Var(X)=E[(X-\mu)^2]$. The variance of a sample gives an estimate of the
    population variance that is biased by a factor of $\frac{n-1}{n}$. The
    best unbiased estimate of the population variance is therefore given by
    $s^2 = \frac{1}{n-1} \sum_{i=1}^n \left(y_i - \overline{y} \right)^2$,
    which is called (unbiased) sample variance.}
}

\newglossaryentry{kurtosis}
{
    name=kurtosis,
    description={ Measure of the peakedness of a distribution, relative to the
    normal distribution.}
}

\newglossaryentry{skewness}
{
    name=skewness,
    description={ Measure of the asymmetry of a distribution.}
}

\newglossaryentry{monteCarlo}
{
    name=Monte Carlo Simulation,
    description={ Repeated simulation of the behavior of some parameter based on repeated
    sampling of a random variable. }
}


\newglossaryentry{markovChain}
{
    name=Markov Chain Model,
    description={ Stochastic model of a process where the probability of each state only
    depends on the previous state.}
}

\newglossaryentry{sensitivity}
{
    name=sensitivity,
    description={ Proportion of actual positives which are correctly identified as such
    (e.g., the percentage of sick people who are correctly identified as
    having the condition).}
}

\newglossaryentry{specificity}
{
    name=specificity,
    description={ Proportion of negatives which are correctly identified as such (e.g.,
     the percentage of healthy people who are correctly identified as not
     having the condition).}
}

\newglossaryentry{hypothesistest}
{
    name=hypothesis test,
    description={ A method of statistical inference used for testing a statistical
    hypothesis.}
}

\newglossaryentry{TypeI}
{
    name=type I error,
    description={ A type I error occurs when one rejects the null hypothesis when it is
    true. The probability of a type I error is the level of significance of
    the test of hypothesis, and is denoted by $\alpha$.}
}

\newglossaryentry{TypeII}
{
    name=type II error,
    description={ A type II error occurs when one rejects the alternative hypothesis (fails
    to reject the null hypothesis) when the alternative hypothesis is true.
    The probability of a type II error is denoted by $\beta$.}
}

\newglossaryentry{power}
{
    name=power,
    description={ Same as Sensitivity. Denoted by $1-\beta$.}
}

\newglossaryentry{paired}
{
    name=paired data,
    description={Two data sets are "paired" when the following one-to-one
    relationship exists between values in the two data sets:
        Each data set has the same number of data points.
        Each data point in one data set is related to one, and only one, data
        point in the other data set.}
}

\newglossaryentry{unpaired}
{
    name=unpaired data,
    description={ Two sets of independent data.}
}

\newglossaryentry{dof_}
{
    name=degrees of freedom,
    description={ The number of degrees of freedom is the number of values in the final
    calculation of a statistic that are free to vary.}
}

\newglossaryentry{correlation}
{
    name=correlation,
    description={ Any departure of two or more random variables from
    independence,}
}

\newglossaryentry{linreg}
{
    name=linear regression,
    description={ Modelling a scalar variable ("dependent variable") using linear predictor
    function. The unknown model parameters are estimated from the data.
    Simple linear regression: $y = k*x + d$.
    Multiple linear regression: $y = k_1*x_1 + k_2*x_2 + ... + k_n*x_n + d$}
}

\newglossaryentry{maxLikelihood}
{
    name=maximum likelihood,
    description={ For a fixed set of data and underlying statistical model, the
    method of maximum likelihood selects the set of values of the model
    parameters that maximizes the likelihood function. Intuitively, this
    maximizes the "agreement" of the selected model with the observed data,
    and for discrete random variables it indeed maximizes the probability of
    the observed data under the resulting distribution. }
}

\newglossaryentry{residual}
{
    name=residual,
    description={ Difference between the observed value and the estimated
    function value.}
}

\newglossaryentry{logreg}
{
    name=logistic regression,
    description={ Also called "logit regression".The probabilities describing the possible
    outcomes of a single trial are modeled, as a function of the explanatory
    (predictor) variables, using a logistic function: $f(x) = \frac{L}{1 +
    \mathrm e^{-k(x-x_0)}}$.}
}

\newacronym{iqr}{IQR}{Inter-Quartile-Rang}

\newacronym{ide}{IDE}{Integrated Development Environment}

\newacronym{pdf}{PDF}{Probability Density Function}

\newacronym{cdf}{CDF}{Cumulative Distribution Function}

\newacronym{sf}{SF}{Survival Function}

\newacronym{isf}{ISF}{Inverse Survival Function}

\newacronym{ppf}{PPF}{Percentile Point Function}

\newacronym{rvs}{RVS}{Random Variate Sample}

\newacronym{kde}{KDE}{Kernel Density Estimation}

\newacronym{dof}{DOF}{Degrees of Freedom}

\newacronym{ci}{CI}{Confidence Interval}

\newacronym{qq-plot}{QQ-Plot}{Quantile-Quantile Plot}

\newacronym{roc}{ROC}{Reiciver Operating Characteristic}

\newacronym{anova}{ANOVA}{ANalysis Of VAriance}

\newacronym{ss}{SS}{Sum of Squares}

\newacronym{hsd}{Tukey HSD}{Tukey Honest Significant Difference test}

\newacronym{mcmc}{MCMC}{Markov Chain Monte Carlo}
